---
title: "Michelson Speed-of-light Measurements"
author: "Lili Baker"
date: 2023-02-02
output:
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
---

*Purpose*: When studying physical problems, there is an important distinction between *error* and *uncertainty*. The primary purpose of this challenge is to dip our toes into these factors by analyzing a real dataset.

*Reading*: [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (Optional)

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to measure the speed of light. His approach was a development upon the method of Foucault[3], and resulted in a new estimate of $v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this challenge, you will analyze Michelson's original data, and explore some of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the value `LIGHTSPEED_VACUUM`) because the meter is actually [*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition) in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse
```

*Data dictionary*:

- `Date`: Date of measurement
- `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
- `Temp`: Ambient temperature (Fahrenheit)
- `Velocity`: Measured speed of light (km / s)

### __q1__ Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n  | MeanVelocity |
|--------------|----|----------|
|            3 | 46 |   299860 |
|            2 | 39 |   299860 |
|            1 | 15 |   299810 |

```{r q1-task}
## TODO: Compute summaries
df_q1 <- df_michelson %>% 
  group_by(Distinctness) %>% 
  summarize(
    n = n(),
    MeanVelocity = mean(Velocity)
  )

df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:
- Write your observations here!
  - The values for the mean velocity across all distinctness values differ from the values Michelson reported. With the exception of the distinctness value of 1, all of the mean velocity values are less than those obtained by Michelson.
- Why might your table differ from Michelson's?
  - The values present in the table may differ from Michelson's as a result of the use of rounding in calculations throughout the process.

The `Velocity` values in the dataset are the speed of light *in air*; Michelson
introduced a couple of adjustments to estimate the speed of light in a vacuum.
In total, he added $+92$ km/s to his mean estimate for `VelocityVacuum` (from
Michelson (1880), pg. 141). While the following isn't fully rigorous ($+92$ km/s
is based on the mean temperature), we'll simply apply this correction to all the
observations in the dataset.

### __q2__ Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- df_michelson %>% 
  mutate(
    VelocityVacuum = Velocity + 92
  )

df_q2
```

As part of his study, Michelson assessed the various potential sources of error,
and provided his best-guess for the error in his speed-of-light estimate. These
values are provided in `LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently, Michelson
believed the true value of the speed-of-light probably lay between
`LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and `LIGHTSPEED_MICHELSON +
LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

- **Error** is the difference between a true value and an estimate of that
  value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
- **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does not know
the error. The best they can do is quantify their degree of uncertainty. We will
learn some means of quantifying uncertainty in this class, but for many real
problems uncertainty includes some amount of human judgment.[2]

### __q3__ Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON
LIGHTSPEED_PM

```

**Observations**:
- Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?
  - Michelson's estimate of the error is less than the true error as he believed that the true value of the speed-of-light would be at most 51 m/s away from his best estimate, while in reality the true value was 151.542 m/s away.
- Make a quantitative comparison between Michelson's uncertainty and his error.
  - Michelson's uncertainty was that the true value of the speed-of-light would lie either 51 m/s less or greater than his estimate. However, when determining the actual error by subtracting the speed-of-light by the true value, it is evident that Michelson underestimated his uncertainty as the actual error was 151.542 m/s, meaning that there is a 101.542 m/s discrepancy between his uncertainty and the actual error exhibited.

The following plot shows all of Michelson's data as a [control chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot is common in manufacturing, where it is used to help determine if a manufacturing process is under [statistical control](https://en.wikipedia.org/wiki/Statistical_process_control). Each dot is one of Michelson's measurements, and the grey line connects the mean taken for each day. The same plot also shows simulated data using a probability model. We'll get into statistics later in the course; for now, let's focus on understanding what real and simulated data tend to look like.

### __q4__ Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:
Similarities
- The data appears to be centered around the same range of values. Both the real and simulated have a wide range of observed values for each day, and these ranges tend to exceed the margins of error.
Differences
- The mean velocities of the real data tend to go outwith the margin of error earlier in the measurement dates, whereas the simulated tends to go out with the bounds later on in the dates of measurement. Additionally, the simulated data tends to vary more than the real data; the real tends to appear in clusters at each day, as opposed to the widespread of measurements seen in the simulated data.

### __q5__ You have access to a few other variables. Construct a few visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?
```{r}

## Visualize
set.seed(101)
df_q2 %>%

  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%


  ggplot(aes(Temp, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dashed",
    color = "red",
  ) +


  geom_line(
    data = . %>%
      group_by(Temp) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  theme_minimal() +
  labs(
    x = "Ambient Temperature (Fahrenheit)",
    y = "Velocity (in Vacuum)"
  )

```
```{r}

## Visualize
set.seed(101)
df_q2 %>%

  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%


  ggplot(aes(Distinctness, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dashed",
    color = "red",
  ) +

  geom_point(
    data = . %>%
      group_by(Distinctness) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "red"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  theme_minimal() +
  labs(
    x = "Distinctness",
    y = "Velocity (in Vacuum)"
  )

```
**Observations**:

- First graph: The real data exhibits a positive trend as the temperature increases. While the mean velocities observed at each temperature do not increase each time the temperature increases, the overall trend of the mean velocity line exhibits a positive increase as the temperature increases. This behavior suggests that the ambient temperature has an impact on the speed of light.
- Second graph: The red points represent the mean velocity values at each level of distinctness. The graph illustrates an increase in the mean velocity values as the distinctness level increases. That being said, the means of distinctness levels of 1 and 2 could be skewed by outliers that fall below the actual speed of light in a vacuum. In that regard, the data recorded with a distinctness level of 3 seems to have a larger quantity of values that fall closer to the actual speed of light, implying that the distinctness level does have an impact on the recorded speed of light.



## Bibliography

- [1] Michelson, [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (1880)
- [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf) (1986)
- [3] BYU video about a [Fizeau-Foucault apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to what Michelson used.
